Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2023-09-25 manuscript/manuscript.Rmd
results and disucssion updates.

There is a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
There is much less agreement regarding the less heavily worked and formally standardized LCTs typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018].
Such forms continue to occur with variable frequency in later time periods [@mcnabb2015], and may be especially prevalent in eastern Asia [@li2021].
Although formal types have been recognized in the Early Acheulean and are commonly used to describe assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018]including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of Early Acheulean large core reduction.

@@

A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance. Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance. Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness). This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p < 0.001, Figure 3c). Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation. A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance. This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; Tables 1 & 2) whereas 3DGM-PC2 more purely isolates convergence (Table 2). The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness. Including a third dimension in the GM analysis generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by both the 2D and 3D analyses remains similar. Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2, cumulatively capturing whether the items are broad and flat or thick and pointed.

@@

@@

We first conducted a stepwise DFA on all flaked pieces (n=192) with inferred technological Mode (one vs. two) as the grouping variable and PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables. All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p< 0.001) which correctly classified 93.8% of artifacts. We thus accepted the validity of classification by Mode in our sample and employed this distinction in subsequent analyses. There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs. Acheulean (n=39) contexts (, p = 0.746). Mode 1 cores from Oldowan contexts (n=37) do include 10 (27%) small, retouched flakes. Only one retouched flake from an Acheulean context was classified by the DFA as a Mode 1 core. This piece, typologically classified as a “knife”, is the smallest (93mm) retouched flake in the Acheulean sample. When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Achuelean contexts. Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs. 186g, Cohen’s d=1.137, p<0.001) and less heavily reduced (mean cSDI 0.057 vs. 0.103, Cohen’s d = -0.884, p < 0.001) despite having similar FAI (mean 0.52 vs. 0.46, Cohen’s d=0.271, p=0.266).

Next, we conducted a stepwise DFA on all flaked Mode 2 pieces (i.e. excluding unmodified large flakes, n = 115) with typology (handaxe, pick, knife) as the grouping variable and the same four independent variables. Both shape PCs and FAI were retained, while cSDI was not entered. This produced two DFs (DF1: Eigenvalue=1.536, 91.3% of variance; DF2: Eigenvalue = 0.146, 8.7% of variance; Wilks Lambda = 0.344; p<0.001) which correctly classified 71.3% of artifacts. Inspection of DF coefficients shows that DF1 captures an inverse correlation between flaked area and pointedness (Linear-PC2) whereas DF2 captures a positive relationship between flaked area and elongation (Linear-PC2). A bivariate plot (Figure 5) illustrates the fact that DF1 captures a range of variation from pointed, heavily-worked picks, through handaxes, to knives, with substantial overlap between adjacent types (Table 2). As an intermediate type, “handaxes” were correctly classified only 43.9% of the time. In agreement with others (Duke, et al., 2021, Kuhn, 2021, Presnyakova, et al., 2018) we conclude that these typological labels artificially partition a continuum of variation and abandon them in subsequent analyses.

##  Effects of Reduction on Shape
To assess the influence of flake removals on core form, we examined the association between our reduction measures (cSDI and FAI) and core shape (PC1, PC2). In the complete sample of flaked pieces (n=192), we observed weak but significant effects of cSDI (r=-0.294, p <0.001) and FAI (r=-0.294, p<0.001) on PC1(flatness) and of FAI only on PC2(pointedness) (r=-0.436, p<0.001). However, it is clear that these overall effects conflate different trends in Mode 1 vs. Mode 2 cores, as well as in cores executed on flake vs. pebble/cobble bases. Within categories, we observed no significant effects of cSDI whereas FAI had variable relationships with core form.

In Mode 1 cores, a weak negative effect of FAI on PC2(pointedness) (n = 76, $\beta$ =-0.006, r=0.240, p= 0.037) suggests that increasing extent of modification tends to slightly increase pointedness (Figure 6a). No effects of FAI on PC1 (flatness) or of SDIc on either PC approach significance. FAI and SDIc are moderately correlated ($\beta$=0.001, r=0.524, p<0.001), an effect that reflects increases in the upper limit of scar density values as flaked area increases (i.e., flaked area constrains maximum cSDI: Figure 6b).

In Mode 2 cores (n=115), reduction measures have different effects depending on initial blank form. In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (Figure 7c,d). Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406). Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p < 0.001). Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e. both upper and lower limits of cSDI increase with FAI).
Comparison of Mode 2 flake cores with unmodified large flakes from Achuelean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1(flatness) and PC2(pointedness). In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs. 0.774 and 0.693 vs. 0.594, respectively) and significantly negative slopes as FAI increases (Figure 9).

A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined. These heavily worked “indeterminate” cores tend to show similar PC1(flatness) and PC2(pointedness) values to cobble-base cores (i.e. thicker and more pointed than flake cores; Figure x). They also appear to follow general trends of shape change with increasing FAI observed across base types (Figure 10).

>>>>

There is a broad consensus that refined handaxes and cleavers from the later Acheulean resulted from procedurally elaborate, skill intensive, and socially learned production strategies [@caruana2020; @garcía-medrano2019; @moore2020; @sharon2009; @shiptonEvolutionSocialTransmission2019; @stout2014] although debate over the presence of explicit, culturally transmitted shape preferences continues [@iovita2011; @moore2020; @shipton2020; @wynn2018].
There is much less agreement regarding the less heavily worked and formally standardized LCTs typical of the earliest Acheulean [@beyene2013; @delatorre2018; @diez-martín2015; @lepre2011; @semaw2018].
Such forms continue to occur with variable frequency in later time periods [@mcnabb2015], and may be especially prevalent in eastern Asia [@li2021].
Although formal types have been recognized in the Early Acheulean and are commonly used to describe assemblages, many workers now see a continuum of morphological variation [@duke2021; @kuhn2020; @presnyakova2018] including the possibility that simple flake production remained an important [@shea2010] or even primary [@moore2016] purpose of Early Acheulean large core reduction.

@@

A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This A PCA on GM transformed linear measures the 33 artifacts identified two PCs accounting for 75.6% of the variance.
Linear-PC1 (Eigenvalue = 0.189) explained 55.2% of the variance.
Factor loadings (**Table** \@ref(tab:tab1)) for Linear-PC1 reflect artifact elongation (i.e., an anti-correlation of length vs. distal width and thickness).
This closely parallels the length vs. width and thickness tradeoff captured by 3DGM-PC1 (Figure 3a) and is reflected in a tight correlation of artifact scores produced by the two PCs (r = 0.903, p \< 0.001, Figure 3c).
Comparison with conventional shape ratios (Table 2) similarly indicates that both Linear-PC1 and 3DGM-PC1 largely capture variation in artifact elongation.
A second factor (Linear-PC2, eigenvalue = 0.07) explained an additional 20.4% of variance.
This factor was less strongly correlated with its 3DGM counterpart (3DGM-PC2; r = 0.344, p = 0.050) probably because Linear-PC2 describes anticorrelated variation in width and thickness (i.e., broad and flat vs. thick and pointed; Table 1) whereas 3DGM-PC2 more purely isolates convergence (Figure 3a).
The remainder of the shape variability explained by Linear-PC2 is captured by higher order 3DGM-PCs like 3DGM-PC3 to 5, which comprise the contribution of the left and right lateral margins to relative thickness.
Use of high-resolution, coordinate-based scan data thus generates PCs that identify more specific shape attributes, but the underlying morphological variability captured by the linear and 3D analyses remains similar.
Together, 3DGM-PC2 (r = 0.344, p = 0.050), 3DGM-PC3 (r = -0.416, p = 0.016), 3DGM-PC4 (r = 0.458, p = 0.007), and 3DGM-PC5 (r = -0.352, p = 0.044) correlate well with Linear-PC2 (Figure 3c), cumulatively capturing whether the items are broad and flat or thick and pointed.
A stepwise regression (r2=0.625, F(4,28)=11.697, p\<0.001, Probability-of-F-to-enter \<= 0.050; Probability-of-F-to-remove \>= 0.100) with Linear-PC2 as the dependent variable retained all four of these 3DGM-PCs as significant predictors.

@@

We thus concluded that linear measures are adequate to capture relevant shape variation and proceeded with a PCA on our full sample.
We identified two PCs accounting for 80.0% of the variance.
PC1 (Eigenvalue = 0.216) explained 56.4% of the variance.
Factor loadings (Table 2) for PC1 reflect artifact flatness (i.e., an anti-correlation of length and width vs. thickness) such that higher values indicate relatively thinner pieces.
PC2 (Eigenvalue = 0.090) explained 23.6% of the variance.
Factor loadings (Table 2) show that PC2 captures artifact convergence or "pointedness" (i.e., an anti-correlation of tip width with length and butt thickness) such that higher values indicate shorter, less pointed forms.

We also tested the validity of our two reduction measures, SDI and FAI.
In agreement with Clarkson (2013), we found that surface area estimated from caliper measures displayed a strong correlation with (r2=0.975, p \< 0.001) but linear over-estimation of ($\beta$ = 1.58) 3DGM surface area (Figure 4a).
This results in a systematic underestimation of SDI that scales with core size (Figure 4b).
However, a simple correction of the caliper estimate (dividing by the slope, 1.58) eliminates surface area over-estimation (mean difference = 256mm2 [\<1.7% of mean], p=0.040) and produces SDI values that agree with 3DGM (r2=0.975, p \< 0.001, $\beta$ = 0.98) (Figure 4c).
We thus proceeded to apply this correction to surface area estimates in the full sample.
Insofar as these relationships are driven by basic geometry, we expect these methods (including correction) to be generalizable to other ESA assemblages.

@@

We first conducted a stepwise DFA on all flaked pieces (n=192) with inferred technological Mode (one vs. two) as the grouping variable and PCs 1 and 2, corrected SDI (cSDI), and FAI as the independent variables.
All variables were retained, yielding one canonical DF (eigenvalue=1.825, Wilks Lambda = 0.354, p\< 0.001) which correctly classified 93.8% of artifacts.
We thus accepted the validity of classification by Mode in our sample and employed this distinction in subsequent analyses.
There was no discernable difference in discriminant scores for Mode 1 cores from Oldowan (n=37) vs. Acheulean (n=39) contexts (, p = 0.746).
Mode 1 cores from Oldowan contexts (n=37) do include 10 (27%) small, retouched flakes.
Only one retouched flake from an Acheulean context was classified by the DFA as a Mode 1 core.
This piece, typologically classified as a "knife", is the smallest (93mm) retouched flake in the Acheulean sample.
When retouched flakes are excluded from the comparison, there are no significant differences in shape between Mode 1 cores from Oldowan and Achuelean contexts.
Interestingly, however, Acheulean Mode 1 cobble-cores are much larger (mean weight 480g vs. 186g, Cohen's d=1.137, p\<0.001) and less heavily reduced (mean cSDI 0.057 vs. 0.103, Cohen's d = -0.884, p \< 0.001) despite having similar FAI (mean 0.52 vs. 0.46, Cohen's d=0.271, p=0.266).

Next, we conducted a stepwise DFA on all flaked Mode 2 pieces (i.e. excluding unmodified large flakes, n = 115) with typology (handaxe, pick, knife) as the grouping variable and the same four independent variables.
Both shape PCs and FAI were retained, while cSDI was not entered.
This produced two DFs (DF1: Eigenvalue=1.536, 91.3% of variance; DF2: Eigenvalue = 0.146, 8.7% of variance; Wilks Lambda = 0.344; p\<0.001) which correctly classified 71.3% of artifacts.
Inspection of DF coefficients shows that DF1 captures an inverse correlation between flaked area and pointedness (Linear-PC2) whereas DF2 captures a positive relationship between flaked area and elongation (Linear-PC2).
A bivariate plot (Figure 5) illustrates the fact that DF1 captures a range of variation from pointed, heavily-worked picks, through handaxes, to knives, with substantial overlap between adjacent types (Table 2).
As an intermediate type, "handaxes" were correctly classified only 43.9% of the time.
In agreement with others [@duke2021; @kuhn2020; @presnyakova2018] we conclude that these typological labels artificially partition a continuum of variation and abandon them in subsequent analyses.

## Effects of Reduction on Shape

To assess the influence of flake removals on core form, we examined the association between our reduction measures (cSDI and FAI) and core shape (PC1, PC2).
In the complete sample of flaked pieces (n=192), we observed weak but significant effects of cSDI (r=-0.294, p \<0.001) and FAI (r=-0.294, p\<0.001) on PC1(flatness) and of FAI only on PC2(pointedness) (r=-0.436, p\<0.001).
However, it is clear that these overall effects conflate different trends in Mode 1 vs. Mode 2 cores, as well as in cores executed on flake vs. pebble/cobble bases.
Within categories, we observed no significant effects of cSDI whereas FAI had variable relationships with core form.

In Mode 1 cores, a weak negative effect of FAI on PC2(pointedness) (n = 76, $\beta$ =-0.006, r=0.240, p= 0.037) suggests that increasing extent of modification tends to slightly increase pointedness (Figure 6a).
No effects of FAI on PC1 (flatness) or of SDIc on either PC approach significance.
FAI and SDIc are moderately correlated ($\beta$=0.001, r=0.524, p\<0.001), an effect that reflects increases in the upper limit of scar density values as flaked area increases (i.e., flaked area constrains maximum cSDI: Figure 6b).

In Mode 2 cores (n=115), reduction measures have different effects depending on initial blank form.
In general, the contrast between flat, lightly-worked flake bases and rounder, more-heavily worked cobble bases tends to inflate relationships between FAI and core shape (Figure 7c,d).
Mode 2 cobble cores (n=37) display a negative effect of FAI on PC1(Flatness) ($\beta$=-0.011, r=0.367, p = 0.026), no effect of FAI on pointedness, and no association between FAI and cSDI (r=0.141, p = 0.406).
Flake cores (n=69) show negative effects of FAI on PC1(Flatness) ($\beta$=-0.007, r=0.284, p = 0.018) and, more strongly, PC2(pointedness) ($\beta$=-0.019, r=0.443, p \< 0.001).
Flake cores show a small but significant positive effect of FAI on cSDI that, in contrast to Mode 1 cores, represents actual covariance rather than simply a relaxation of constraint (i.e. both upper and lower limits of cSDI increase with FAI).

Comparison of Mode 2 flake cores with unmodified large flakes from Achuelean contexts (n=35) shows that ranges of shape variation substantially overlap but that flaking generally reduces both PC1(flatness) and PC2(pointedness).
In fact, regressions of PC1 and PC2 on FAI show y-intercepts that closely approximate unmodified flake mean values (0.852 vs. 0.774 and 0.693 vs. 0.594, respectively) and significantly negative slopes as FAI increases (Figure 9).

A small number of Mode 2 cores (n=9) were so heavily modified (mean FAI=89%, range 65-100) that the blank form could not be determined.
These heavily worked "indeterminate" cores tend to show similar PC1(flatness) and PC2(pointedness) values to cobble-base cores (i.e. thicker and more pointed than flake cores; Figure x).
They also appear to follow general trends of shape change with increasing FAI observed across base types (Figure 10).

# Discussion

Following the criteria we proposed in the Introduction, Large Flake base cores from Acheulean contexts show evidence of design.
Importantly, "design" in this context is understood as "design in four dimensions" sensu Kuhn [-@kuhn2020].
Increasing FAI is associated with progressive narrowing ( i.e., decreasing PC1) and convergence (decreasing PC2) whereas cSDI is not.
This indicates that the size and non-overlapping placement of flake removals drives shape change, rather than simply the removal of mass in general as might be expected from conventional models of resharpening (?).
The fact that cSDI nevertheless correlates with FAI is consistent with relatively light (number of scars per piece ranges from 2 to 17 with a mean of 6.94) and spatially distributed re-touching involving relatively little overlap between scars.
Such a pattern could be produced by intentional imposition of a design target or "mental template" or a more procedural/functional bias in favor of establishing cutting edges at one end of the long axis.
In the absence of effective thinning/volume management techniques, this will lead to rapid narrowing/thickening of the piece overall and especially at the "tip." This is consistent with differences between flaked vs. unmodified large flakes and furthers the point that blank production already can be conceived as a process of "design" -- many unmodified flakes may have been similar in shape and just as useful as many retouched tools.

This contrasts with reduction of Mode 1 cores.
Unexpectedly, FAI did have an effect to increase convergence.
However, this effect was substantially weaker than seen in LFB Acheulean cores and may simply reflect the fact that reduction is unlikely to be perfectly symmetrical and our orienting protocol guarantees that one end is always more pointed than the other unless they are precisely equal in breath.
In support of this "byproduct" interpretation, the relationship between FAI and cSDI is essentially what one would expect from knapping unconstrained by any systematic core-shaping strategy (greater flaked area can, but does not necessarily, accommodate a greater number of flake scars)

Our most complicated results come from Mode 2 cobble cores.
Evidence of intentional shaping is limited: FAI has a moderate effect on PC1(flatness) and no effect on pointedness.
Furthermore, the is no association between FAI and cSDI cobble cores display a greater range of cSDI values.
Instead, FAI is strongly correlated with scar size (mean scar length: n=33, r=.588, P\<0.001; Figure 11).
Together, this suggests that the removal of large flakes in particular increases FAI and tends to narrow the piece.
We propose that this pattern may actually indicate that at least some cobble cores were precisely that: cores for the generation of LFBs.
Given the difficulty of working relatively spherical cobbles, we expect that knappers would have chosen elongated and/or tabular cobbles.
Attempts to maximize potential flake dimensions would naturally lead to removals along the long axis rather than at either end.
This hypothesis was suggested by informal "experiential" knapping by the lead author with Gona cobbles, during which LFB production produced pick like forms as a byproduct.
In support of the plausibility of this hypothesis, the maximum scars length on Gona Mode 2 cobble cores substantially overlaps with the size of unmodified large flakes from Acheulean contexts.
It is also possible that cobble cores might begin as cores for LFB production and then be finished into core tools in their own right.
These possibilities have major implications for understanding the occurrence and apparent morphological stability [@beyene2013] of "picks" throughout the African Acheuelan but remains to be tested.
